---
title: "STAT 665 Homework 2"
author: "Issac Li"
date: "1/30/2017"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Part I

Impute the NAs in  ```capital_run_length_average``` with k-nn algorithm. 

```{r impute nas}
require(FNN)
require(data.table)

train=fread("/Users/lizhuo/Documents/STAT665/HW2/spam_train.csv",header = T)
test=fread("/Users/lizhuo/Documents/STAT665/HW2/spam_test.csv",header = T)
paste(sum(is.na(train$capital_run_length_average))," missing values in Training data")
paste(sum(is.na(test$capital_run_length_average))," missing values in Testing data")
```

The missing values are approximately distributed with similar proportions so we can merge the two datasets and then do imputation
to yield better imputation results. 

```{r impute}
# Combining datasets
train$group=1
test$group=2;test$spam=NA
merged=rbind(train,test)
# Rescaling and add group numbers for reverse mapping
merged=data.table(cbind(merged$group,merged$spam,merged$capital_run_length_average,apply(merged[,!c("group","spam","capital_run_length_average"),with=F],2,function(x) scale(x,center = min(x),scale=max(x)-min(x)))))

colnames(merged)[1:3]<-c("group","spam","capital_run_length_average")

# Prepare dataset for imputation
impute_test=merged[is.na(merged$capital_run_length_average)]
impute_train=merged[!is.na(merged$capital_run_length_average)]

impute_test_x=impute_test[,!c("capital_run_length_average"),with=F]

impute_train_x=impute_train[,!c("capital_run_length_average"),with=F]
impute_train_y=impute_train$capital_run_length_average

ans=knn.reg(train =impute_train_x[,!c("group","spam"),with=F],test = impute_test_x[,!c("group","spam"),with=F], y = impute_train_y,k=15)

impute_test$capital_run_length_average=ans$pred

merged=rbind(impute_test,impute_train)

ctrain=merged[group==1,!c("group"),with=F]
ctest=merged[group==2,!c("group","spam"),with=F]
  
paste(sum(is.na(ctrain$capital_run_length_average))," missing values in Training data")
paste(sum(is.na(ctest$capital_run_length_average))," missing values in Testing data")
```
After imputation we can see that there is no more NA’s in the ```capital_run_length_average```
column in either the training or the test set


## Part II
#### Please see HW2_knnclass.R
```{r KNN class, echo=FALSE}
knnclass <- function(xtrain, xtest, ytrain) {
  # Standarize columns
  mu = colMeans(xtrain)
  sd = sqrt(apply(xtrain,2,var))
  strain = data.table(apply(xtrain,2,function(x) (x-mean(x))/sqrt(var(x))))
  # Split training dataset
  train_ind <- sample(seq_len(nrow(xtrain)), size = floor(nrow(xtrain)*0.8))
  subxtrain = xtrain[train_ind,]
  subxtest = xtrain[-train_ind,]
  subytrain = ytrain[train_ind]
  subytest = ytrain[-train_ind]
  
  # Select optimal k
  subytrain=as.numeric(as.matrix(subytrain))
  cal_d<-function(xvec1,xmatrix){
    apply(xmatrix,1,function (x) sqrt(sum((xvec1-x)^2)))
  }
  dmatrix=apply(subxtest,1,cal_d,xmatrix=subxtrain)

  cal_class<-function(subxtest,dmatrix,subytrain,k=3){
    kmatrix=apply(t(dmatrix),1,order)[2:k+1,]
    yresponse=colMeans(apply(kmatrix,2,function(x) subytrain[x]))
    yclass=ifelse(yresponse>=0.5,1,0)
    return(yclass)
  }
  
  MSE_k=c()
  for (k in seq(3,nrow(subxtest)/10)){
  yclass=cal_class(subxtest,dmatrix,subytrain,k=k)
  MSE_k=c(MSE_k,mean(sum((yclass-subytest)^2)))
  }
  
  op_k=(which.min(MSE_k)+3)
  dmatrix=apply(xtest,1,cal_d,xmatrix=xtrain)
  yclass=cal_class(xtest,dmatrix,ytrain,k=op_k)

return(yclass)
}
```

## Part III
```{r p3,warning=F,message=F}
# Use KNN to Predict spam without ```capital_run_length_average```
knn_pred1 = knnclass(xtrain = ctrain[,!c("capital_run_length_average","spam"),with=F],
                     xtest = ctest[,!c("capital_run_length_average"),with=F],
                     ytrain = ctrain$spam)


# Use KNN to Predict spam with ```capital_run_length_average```
knn_pred2 = knnclass(xtrain = ctrain[,!c("spam"),with=F],
                     xtest = ctest,
                     ytrain = ctrain$spam)

# Use logistic regression to predict spam without ```capital_run_length_average```
log_fit1=glm(data=ctrain,formula = spam ~ .-capital_run_length_average,family = "binomial")
logm_pred1=predict(log_fit1,newdata = ctest,type = "response")
logm_pred1=ifelse(logm_pred1>=0.5,1,0)

# Use logistic regression to predict spam with ```capital_run_length_average```
log_fit2=glm(data=ctrain,formula = spam ~ .,family = "binomial")
logm_pred2=predict(log_fit2,newdata = ctest,type = "response")
logm_pred2=ifelse(logm_pred2>=0.5,1,0)

summary.glm(log_fit2)
results=cbind(ctest$capital_run_length_average,knn_pred1,knn_pred2,logm_pred1,logm_pred2)
colnames(results)[1]<-"capital_run_length_average"
write.csv(results,file = "HW2_zl368_results.csv")
```

Based on the regression summary, we can see that emails with the following characteristics correlate with high probability of being spam:    
  1. High frequence of the following words:  
    * our, over, remove, internet, mail, free, business, your, 000, money, hp, hpl,george, 85, technology, pm, meeting, confernce
  2. High frequence of the following chars:  
    * $, !
  3. Long capital run length:
    * Length of longest capital run length > total capital run length > average capital run length



